Post-Disturbance Tree Community Trajectories in a Neotropical Forest

### 2018-07-23 WholePlotTrajectories.Rmd
coquilles

The long-term impact of disturbance on community functional redundancy meant a lower resilience of the pre-disturbance communities, with higher chances to see the persistence of disturbance-specific species at the expense of late-successional ones [@Haddad2008].
Besides, the long-term recovery of infrequent species increases the risks to loose cornerstone species, with unexpected ecological consequences [@Jones1994;@Chazdon2003a;@Diaz2005].
Apart from the functional characteristics considered here, infrequent species might indeed have unique functions in the ecosystem or be a key for some fauna.

>>>>

The long-term impact of disturbance on community functional redundancy meant a lower resilience of the pre-disturbance communities, with higher chances to see the persistence of disturbance-specific species at the expense of late-successional ones [@Haddad2008].
Besides, the long-term recovery of infrequent species increases the risks to loose cornerstone species, with unexpected ecological consequences [@Jones1994;@Chazdon2003a;@Diaz2005].
Apart from the functional characteristics considered here, infrequent species might indeed have unique functions in the ecosystem or be a key for some fauna [@Schleuning2016].

### 2018-07-23 WholePlotTrajectories.tex
coquilles

In tropical forests, ecological communities are regularly re-shaped by
natural disturbance events changing both the abiotic environment,
through the fluxes of light, heat and water \citep{Goulamoussene2017},
and the biotic interactions such as competition among species
\citep{Chesson2000, Herault2018}. One of the cornerstone of tropical
forest ecology is to understand the processes and drivers of ecosystems
response to disturbance \citep{Chazdon2003a}. For now, this has been
largely studied through forest structural parameters such as aboveground
biomass, tree height or stem density
\citep{Piponiot2016, Rutishauser2016} that are rapid and convenient to
measure. These structural parameters have been sucessfully modeled,
giving important insights into the recovery of ecosystem processes and
services \citep{Herault2018}. However the response of forests diversity
and composition is still unclear, albeit it determines the productivity,
stability and functioning of ecosystems \citep{Tilman2014, Liang2016}.
In the short-term, moderate disturbance may lead to positive impacts on
communities diversity, an idea formalized by the intermediate
disturbance hypothesis (IDH) stating a maximized species diversity when
disturbance intensity is not too high \citep{Molino2001, Kariuki2006a}.

@@

This variability of protocols in time raised methodological issues as
vernacular names usually correspond to different botanical species. It
resulted in significant taxonomic uncertainties that had to be
propagated to composition and diversity metrics. The uncertainty
propagation was done through a Bayesian framework reconstituting
complete inventories at genus level from real incomplete ones on the
basis of vernacular/botanical names association. Vernacular names were
replaced through multinomial trials based on the association probability
\(\big[\alpha_1, \alpha_2,..., \alpha_3\big]\) observed across all
inventories between each vernacular name \emph{v} and all species
\(\big[s_1, s_2,..., s_N\big]\):

\begin{align}
M_v\Big(\big[s_1, s_2,..., s_N\big],\big[\alpha_1, \alpha_2,..., \alpha_3\big]\Big) \nonumber
\end{align}

@@

While both taxonomic and functional composition remained stable in
undisturbed communities (Figure \ref{fig:NMDSplans}), they followed
marked and consistent trajectories over post-disturbance time. In
disturbed communities, these compositional changes corresponded to
shifts towards species with more acquisitive functional strategies, from
communities with high average WSG to high average SLA and chlorophyll
content (see appendix I). For functional composition, this translated
into cyclic compositional changes with an unachieved recovery of the
initial composition (Figure \ref{fig:NMDSplans}). The maximum
dissimilarity with the initial state was positively correlated to the
disturbance intensity for both taxonomic and functional composition
(\(\rho_{spearman}^{taxonomic}=0.87\) and
\(\rho_{spearman}^{functional}=0.90\) respectively). The maximum value
was reached around 26 years after disturbance for taxonomic composition
and 22 years for functional composition.

@@

The long-term impact of disturbance on community functional redundancy
meant a lower resilience of the pre-disturbance communities, with higher
chances to see the persistence of disturbance-specific species at the
expense of late-successional ones \citep{Haddad2008}. Besides the
long-term recovery of infrequent species increases the risks to loose
cornerstone species, with unexpected ecological consequences
\citep{Jones1994, Chazdon2003a, Diaz2005}. Apart from the functional
characteristics considered here, infrequent species might indeed have
unique functions in the ecosystem or be a key for some fauna.

@@

Our study revealed community recovery through the combination of
deterministic processes driving their convergence in functional
composition, and dispersal limitation maintaining their divergence in
taxonomic composition. The IDH was validated for community taxonomic
richness and, to some extent, taxonomic evenness but disproved regarding
community functional richness and evennes that were enhanced for any
disturbance intensity by the high functional differences of pioneers
compared to late-successional functoinal composition. The IDH was
translated in time by the recruitment, beyond an intensity threshold, of
short-lived pionneers that prevented in the first times the
establishment of more diverse long-lived pionneers, recruited otherwise
below the intensity threshold. The resilience of tropical forests proved
consistent although several decades-long. Still, the disturbance impact
on communities redundancy cautioned against the risks of infrequent
species loss and the persistence of disturbance-specific communities
\citep{Gourlet-Fleury2005}. As the trajectories highlighted the
recruitment processes proved central for communities response to
disturbance and closer focus on demographical drivers of communities
response would clarify the fate of the future forests.

>>>>

In tropical forests, ecological communities are regularly re-shaped by
natural disturbance events changing both the abiotic environment,
through the fluxes of light, heat and water \citep{Goulamoussene2017},
and the biotic interactions such as competition among species
\citep{Chesson2000, Herault2018}. One of the cornerstone of tropical
forest ecology is to understand the processes and drivers of ecosystems
response to disturbance \citep{Chazdon2003a}. For now, this has been
largely studied through forest structural parameters such as aboveground
biomass, tree height or stem density
\citep{Piponiot2016, Rutishauser2016} that are rapid and convenient to
measure. These structural parameters have been sucessfully modeled,
giving important insights into the recovery of ecosystem processes and
services \citep{Herault2018}. However the response of forests diversity
and composition remains unclear, albeit it determines the productivity,
stability and functioning of ecosystems \citep{Tilman2014, Liang2016}.
In the short-term, moderate disturbance may lead to positive impacts on
communities diversity, an idea formalized by the intermediate
disturbance hypothesis (IDH) stating a maximized species diversity when
disturbance intensity is not too high \citep{Molino2001, Kariuki2006a}.

@@

This variability of protocols in time raised methodological issues as
vernacular names usually correspond to different botanical species. It
resulted in significant taxonomic uncertainties that had to be
propagated to composition and diversity metrics. The uncertainty
propagation was done through a Bayesian framework reconstituting
complete inventories at genus level from real incomplete ones on the
basis of vernacular/botanical names association. Vernacular names were
replaced through multinomial trials based on the association probability
\(\big[\alpha_1, \alpha_2,..., \alpha_V\big]\) observed across all
inventories between each vernacular name \emph{v} and all species
\(\big[s_1, s_2,..., s_N\big]\):

\begin{align}
M_v\Big(\big[s_1, s_2,..., s_N\big],\big[\alpha_1, \alpha_2,..., \alpha_V\big]\Big) \nonumber
\end{align}

@@

While both taxonomic and functional composition remained stable in
undisturbed communities (Figure \ref{fig:NMDSplans}), they followed
marked and consistent trajectories over post-\break disturbance time. In
disturbed communities, these compositional changes corresponded to
shifts towards species with more acquisitive functional strategies, from
communities with high average WSG to high average SLA and chlorophyll
content (see appendix I). For functional composition, this translated
into cyclic compositional changes with an unachieved recovery of the
initial composition (Figure \ref{fig:NMDSplans}). The maximum
dissimilarity with the initial state was positively correlated to the
disturbance intensity for both taxonomic and functional composition
(\(\rho_{spearman}^{taxonomic}=0.87\) and
\(\rho_{spearman}^{functional}=0.90\) respectively). The maximum value
was reached around 26 years after disturbance for taxonomic composition
and 22 years for functional composition.

@@

The long-term impact of disturbance on community functional redundancy
meant a lower resilience of the pre-disturbance communities, with higher
chances to see the persistence of disturbance-specific species at the
expense of late-successional ones \citep{Haddad2008}. Besides, the
long-term recovery of infrequent species increases the risks to loose
cornerstone species, with unexpected ecological consequences
\citep{Jones1994, Chazdon2003a, Diaz2005}. Apart from the functional
characteristics considered here, infrequent species might indeed have
unique functions in the ecosystem or be a key for some fauna
\citep{Schleuning2016}.

@@

Our study revealed community recovery through the combination of
deterministic processes driving their convergence in functional
composition, and dispersal limitation maintaining their divergence in
taxonomic composition. The IDH was validated for community taxonomic
richness and, to some extent, taxonomic evenness but disproved regarding
community functional richness and evennes that were enhanced for any
disturbance intensity by the high functional differences of pioneers
compared to late-successional functoinal composition. The IDH was
translated in time by the recruitment, beyond an intensity threshold, of
short-lived pionneers that prevented in the first times the
establishment of more diverse long-lived pionneers, recruited otherwise
below the intensity threshold. The resilience of tropical forests proved
consistent although several decades-long. Still, the disturbance impact
on communities redundancy cautioned against the risks of infrequent
species loss and the persistence of disturbance-specific communities
\citep{Gourlet-Fleury2005}.
